---
title: "Untitled"
format: html
editor: visual
---

```{r}
library(terra)
library(rasterVis)
library(ggmap)
library(tidyverse)
library(knitr)
library(sf)
# New Packages
library(ncdf4) # to import data from netcdf format
```

```{r}
# Create afolder to hold the downloaded data
dir.create("data",showWarnings = F) #create a folder to hold the data

lulc_url="https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MCD12Q1.051_aid0001.nc?raw=true" #landcover

lst_url="https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MOD11A2.006_aid0001.nc?raw=true" #land surface temperature

# download them
download.file(lulc_url,destfile="data/MCD12Q1.051_aid0001.nc", mode="wb")
download.file(lst_url,destfile="data/MOD11A2.006_aid0001.nc", mode="wb")
```
# Load data
```{r}
lulc=rast("data/MCD12Q1.051_aid0001.nc",subds="Land_Cover_Type_1")
# land surface temperature
lst=rast("data/MOD11A2.006_aid0001.nc",subds="LST_Day_1km") #land cover
# Explore data 
# Land Surface Temperature 
# Plot lulc data 
plot(lulc)

# Convert LST to Degree C
lulc=lulc[[13]]
plot(lulc)
```

```{r}
### Process landcover data
Land_Cover_Type_1 = c(
    Water = 0, 
    `Evergreen Needleleaf forest` = 1, 
    `Evergreen Broadleaf forest` = 2,
    `Deciduous Needleleaf forest` = 3, 
    `Deciduous Broadleaf forest` = 4,
    `Mixed forest` = 5, 
    `Closed shrublands` = 6,
    `Open shrublands` = 7,
    `Woody savannas` = 8, 
    Savannas = 9,
    Grasslands = 10,
    `Permanent wetlands` = 11, 
    Croplands = 12,
    `Urban & built-up` = 13,
    `Cropland/Natural vegetation mosaic` = 14, 
    `Snow & ice` = 15,
    `Barren/Sparsely vegetated` = 16, 
    Unclassified = 254,
    NoDataFill = 255)

lcd=data.frame(
  ID=Land_Cover_Type_1,
  landcover=names(Land_Cover_Type_1),
  col=c("#000080","#008000","#00FF00", "#99CC00","#99FF99", "#339966", "#993366", "#FFCC99", 
        "#CCFFCC", "#FFCC00", "#FF9900", "#006699", "#FFFF00", "#FF0000", "#999966", "#FFFFFF", 
        "#808080", "#000000", "#000000"),
  stringsAsFactors = F)
# colors from https://lpdaac.usgs.gov/about/news_archive/modisterra_land_cover_types_yearly_l3_global_005deg_cmg_mod12c1kable(head(lcd))
```

```{r}
# convert to raster (easy)lulc=as.factor(lulc)

# update the RAT with a left join

#levels(lulc)=left_join(levels(lulc)[[1]],lcd)[-1,]#levels():return raster RAT for inspections
# join with color table(lcd) by 'ID' column and delete first row (water)activeCat(lulc)=1 # get or set the active category

# plot it
gplot(lulc)+
  geom_raster(aes(fill=as.factor(value)))+
  scale_fill_manual(values=setNames(lcd$col,lcd$ID),
                    labels=lcd$landcover,
                    breaks=lcd$ID,
                    name="Landcover Type")+
  coord_equal()+
  theme(legend.position = "right")+guides(fill=guide_legend(ncol=1,byrow=TRUE))
```
Convert LULC raster into a ‘factor’ (categorical) raster. This requires building the Raster Attribute Table (RAT). Unfortunately, this is a bit of manual process as follows.

```{r}
# convert to raster (easy)
lulc=as.factor(lulc)

# update the RAT with a left join
levels(lulc)=left_join(levels(lulc)[[1]],lcd)[-1,] # levels(): return raster RAT for inspection 
# join with color table (lcd) by 'ID' column and delete first row (water)
activeCat(lulc)=1 # get or set the active category

# plot it
gplot(lulc)+
  geom_raster(aes(fill=as.factor(value)))+
  scale_fill_manual(values=setNames(lcd$col,lcd$ID),
                    labels=lcd$landcover,
                    breaks=lcd$ID,
                    name="Landcover Type")+
  coord_equal()+
  theme(legend.position = "right")+
  guides(fill=guide_legend(ncol=1,byrow=TRUE))
```
## Land Surface Temperature
```{r}
plot(lst[[1:12]])
```
## Convert LST to Degrees C
You can convert LST from Degrees Kelvin (K) to Celcius (C) with scoff().
```{r}
scoff(lst) = cbind(0.02,-273.15) #scale function
plot(lst[[1:10]])
```
Extract LST values for a single point and plot them.
# Part 1: Extract timeseries for a point
```{r}
# Create a point 
lw= data.frame(x= -78.791547,y=43.007211) %>% st_as_sf(coords=c("x","y"),crs=4326)

# transform the point
lw_rast <- st_transform(lw, crs = st_crs(lst))

# Extract the LST data for that location 
extracted_LST <- terra::extract(lst, lw_rast, buffer=1000, fun=mean, na.rm=T) %>% select(-1) %>% t() %>% 
  as.data.frame()

time(lst)

# combine extracted LST data and time data
LST_time <- cbind.data.frame(extracted_LST, time(lst))
colnames(LST_time) <-  c("mean_temp", "date") # change column name

# visualize 
ggplot(LST_time, aes(x = date, y = mean_temp)) +
  geom_point() +
  geom_line() +
  geom_smooth(span= 0.05) +
  labs(x = 'date', y = 'Monthly Mean Land Surface Temperature')

# Assuming LST_time is your data frame and it has columns 'month' and 'mean_temp'
LST_time <- data.frame(
  month = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"),
  mean_temp = c(18.53, 25.66, 0.18, -2.72, -5.63, 27.22, 25.68, 8.4, 22.61, 8.23, 14.82, 21.73)
)

# Convert month names to date format for proper plotting
LST_time$month <- factor(LST_time$month, levels = month.name)

ggplot(LST_time, aes(x = month, y = mean_temp, group = 1)) +
  geom_point() +
  geom_line() +
  geom_smooth(span = 0.3) +  # Adjusted span for better smoothing
  labs(x = 'Month', y = 'Monthly Mean Land Surface Temperature (°C)') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Part 2: Summarize weekly data to monthly climatologies
```{r}
# summarize the mean value per month
lst_month <- tapp(lst, index='month', fun=mean, na.rm = TRUE) # apply a function to subsets of layer of a SpatRaster

# set names of layers to months 
names(lst_month) =
  month.name[as.numeric(str_replace(names(lst_month),"m_",""))]
3.
# Plot the map for each month with gplot() in the RasterVis Package
gplot(lst_month) + geom_raster(aes(fill = value)) +
          facet_wrap(~ variable) + 
          theme(axis.text.x = element_blank(),
                axis.text.y = element_blank()) +
          scale_fill_gradientn(colors = c("blue", "white", "red")) +
          labs(x = 'x', y = 'y')

# Calculate the monthly mean for the entire image 
global(lst_month,mean,na.rm=T)
```

Part 3: Summarize Land Surface Temperature by Land Cover
```{r}
# Resample lulc to lst grid using resample() with method=near to create a new object called lulc2
lulc2 <- resample(lulc, lst, method = 'near')

# Extract the values from lst_month and lulc2 into a data.frame as follows:
lcds1=cbind.data.frame(
values(lst_month),
ID=values(lulc2[[1]]))%>%
na.omit()

lcds1 %>% 
  
# Gather the data into a ‘tidy’ format 
  gather(key='month',value='value',-Land_Cover_Type_1_13)%>%
  
  # convert ID to numeric 
mutate(ID=as.numeric(Land_Cover_Type_1_13)) %>%
  # left join with lcd (lcd: landcover and color table)
  left_join(lcd, by='ID') %>% 
  # filter 
  filter(landcover%in%c("Urban & built-up","Deciduous Broadleaf forest")) %>%
  # visualize
  ggplot(aes(x = month, y = value), alpha = 0.3) +
    geom_jitter() +
    geom_violin(
      alpha = 0.7, fill='#A4A4A4', color="red") +
    facet_wrap(~landcover, ncol = 2) +
    labs(x = 'Month', y = 'Monthly Mean Land Surface Temperature (C)') +
    ggtitle('Land Surface Temperature in Urban and Forest')
```
```{r}
# Sample data
data <- data.frame(
  month = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"),
  Deciduous_Broadleaf_forest = c("18.53 (±0.82)", "25.66 (±1.1)", "0.18 (±0.51)", "-2.72 (±0.55)", "-5.63 (±0.98)", "27.22 (±1.17)", "25.68 (±1.17)", "8.4 (±0.88)", "22.61 (±1.02)", "8.23 (±0.55)", "14.82 (±0.55)", "21.73 (±0.72)"),
  Urban_built_up = c("20.14 (±3.36)", "29.84 (±2.44)", "0.7 (±0.62)", "-1.3 (±0.86)", "-4.3 (±0.86)", "31.52 (±2.81)", "29.77 (±3.16)", "10.43 (±2.73)", "25.7 (±3.34)", "8.59 (±0.81)", "16.53 (±1.2)", "24.72 (±1.86)")
)

# Gather the data
long_data <- data %>%
  gather(key = "landcover_class", value = "temperature", -month)

# Spread the data
wide_data <- long_data %>%
  separate(temperature, into = c("mean", "sd"), sep = " \\(±") %>%
  mutate(sd = sub("\\)", "", sd)) %>%
  unite("temperature", mean, sd, sep = " (±") %>%
  spread(key = "landcover_class", value = "temperature")

# View the final table
print(wide_data)
```



